---
title: "Create Data - Crime"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates the crime data sets needed for the Spring 2019 edition of the SOC 4650/5650 final projects.

## Dependencies
This notebook requires a number of packages to working with data and wrangling it.

```{r load-packages}
# tidystl packages
library(compstatr)

# tidyverse packages
library(dplyr)
library(purrr)
library(readr)

# other packages
library(here)
library(testthat)
```

## Prepare Data
### Create Objects
First, we prep the raw data by converting filenames like `January2018.CSV.html` to `january2018.csv`. This is done with the `cs_prep_year()` function, which is combined here with `map()` to iterate over a vector of years that correspond to the raw data subfolders.

```{r prep-data}
# create vector of years to clean
years <- as.character(2008:2018)

# clean data files
years %>%
  split(years) %>%
  map(~ cs_prep_year(here("data", "raw", .x))) -> out

# remove output
rm(years, out)
```

### Load Raw Data
Next, we load year-list objects using `cs_load_year()`:

```{r load-data}
data2018_raw <- cs_load_year(here("data", "raw", "2018"))
data2017_raw <- cs_load_year(here("data", "raw", "2017"))
data2016_raw <- cs_load_year(here("data", "raw", "2016"))
data2015_raw <- cs_load_year(here("data", "raw", "2015"))
data2014_raw <- cs_load_year(here("data", "raw", "2014"))
data2013_raw <- cs_load_year(here("data", "raw", "2013"))
```

### 2018
We validate the data to make sure it can be collapsed using `cs_validate_year()`:

```{r validate-data18}
cs_validate_year(data2018_raw, year = "2018")
```

Since the validation result is a value of `TRUE`, we can proceed to collapsing the year-list object into a single tibble with `cs_collapse()` and then stripping out crimes reported in 2018 for earlier years using `cs_combine()`. We also strip out unfounded crimes that remain using `cs_filter_count()`:

```{r collapse-data18}
# collapse into single object
data2018_raw <- cs_collapse(data2018_raw)

# combine and filter
cs_combine(type = "year", date = 2018, data2018_raw) %>%
  cs_filter_count(var = Count) -> data2018
```

The `data2018` object now contains only crimes reported in 2018.

### 2017
We'll repeat the validation process with the 2017 data:

```{r validate-data17}
cs_validate_year(data2017_raw, year = "2017")
```

Since we fail the validation, we can use the `verbose = TRUE` option to get a summary of where validation issues are occuring. 

```{r validate-data17-verbose}
cs_validate_year(data2017_raw, year = "2017", verbose = TRUE)
```

The data for May 2017 do not pass the validation checks. We can extract this month and confirm that there are too many columns in the May 2017 release. Once we have that confirmed, we can standardize that month and re-run our validation.

```{r fix-may17}
# extract data and unit test column numbers
expect_equal(ncol(cs_extract_month(data2017_raw, month = "May")), 26)

# standardize months
data2017_raw <- cs_standardize(data2017_raw, month = "May", config = 26)

# validate data
cs_validate_year(data2017_raw, year = "2017")
```

We now get a `TRUE` value for `cs_validate_year()` and can move on to collapsing the 2017 and 2018 raw data objects to create a new object, `data2017`, that contains all known 2017 crimes including those that were reported or upgraded in 2018.

```{r collapse-data17}
# collapse into single object
data2017_raw <- cs_collapse(data2017_raw)

# combine and filter
cs_combine(type = "year", date = 2017, data2018_raw, data2017_raw) %>%
  cs_filter_count(var = Count) -> data2017
```

### 2016
We'll repeat the validation process with the 2016 data:

```{r validate-data16}
cs_validate_year(data2016_raw, year = "2016")
```

Since the validation process passes, we can immediately move on to creating our 2016 data object:

```{r collapse-data16}
# collapse into single object
data2016_raw <- cs_collapse(data2016_raw)

# combine and filter
cs_combine(type = "year", date = 2016, data2018_raw, data2017_raw, data2016_raw) %>%
  cs_filter_count(var = Count) -> data2016
```

### 2015
We'll repeat the validation process with the 2015 data:

```{r validate-data15}
cs_validate_year(data2015_raw, year = "2015")
```

Since the validation process passes, we can immediately move on to creating our 2015 data object:

```{r collapse-data15}
# collapse into single object
data2015_raw <- cs_collapse(data2015_raw)

# combine and filter
cs_combine(type = "year", date = 2015, data2018_raw, data2017_raw, data2016_raw, data2015_raw) %>%
  cs_filter_count(var = Count) -> data2015
```

### 2014
We'll repeat the validation process with the 2014 data:

```{r validate-data14}
cs_validate_year(data2014_raw, year = "2014")
```

Since the validation process passes, we should be able to immediately move on to creating our 2014 data object. However, we get an error when we go to collapse our data because `ILEADSAddrress` is character in a particular month:

```{r fix-jan14}
# extract data
jan2014 <- cs_extract_month(data2014_raw, month = "January")

# unit test column number
expect_equal(class(jan2014$ILEADSAddress), "character")

# fix ILEADSAddress
jan2014 <- mutate(jan2014, ILEADSAddress = as.numeric(ILEADSAddress))

# replace data
data2014_raw <- cs_replace_month(data2014_raw, month = "January", jan2014)

# remove object
rm(jan2014)

# validate data
cs_validate_year(data2014_raw, year = "2014")
```

After double-checking our validation, we can now collapse our data:

```{r collapse-data14}
# collapse into single object
data2014_raw <- cs_collapse(data2014_raw)

# combine and filter
cs_combine(type = "year", date = 2014, data2018_raw, data2017_raw, data2016_raw, data2015_raw, data2014_raw) %>%
  cs_filter_count(var = Count) -> data2014
```

### 2013
We'll repeat the validation process with the 2013 data:

```{r validate-data13}
cs_validate_year(data2013_raw, year = "2013")
```

Since we fail the validation, we can use the `verbose = TRUE` option to get a summary of where validation issues are occuring. 

```{r validate-data13-verbose}
cs_validate_year(data2013_raw, year = "2013", verbose = TRUE)
```

The data for January through May, July, and August do not pass the validation checks. We can extract these and confirm that there are not enough columns in each of these releases Once we have that confirmed, we can standardize that month and re-run our validation.

```{r fix-month13}
# January - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "January")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "January", config = 18)

# February - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "February")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "February", config = 18)

# March - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "March")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "March", config = 18)

# April - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "April")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "April", config = 18)

# May - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "May")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "May", config = 18)

# July - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "July")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "July", config = 18)

# August - extract data, unit test, and standardize
expect_equal(ncol(cs_extract_month(data2013_raw, month = "August")), 18)
data2013_raw <- cs_standardize(data2013_raw, month = "August", config = 18)
# remove object
rm(month13)

# validate data
cs_validate_year(data2013_raw, year = "2013")
```

We now get a `TRUE` value for `cs_validate_year()` and can move on to collapsing our raw data objects to create a new object, `data2013`, that contains all known 2013 crimes including those that were reported or upgraded in subsequent years:

```{r collapse-data13}
# collapse into single object
data2013_raw <- cs_collapse(data2013_raw)

# combine and filter
cs_combine(type = "year", date = 2013, data2018_raw, data2017_raw, data2016_raw, data2015_raw, data2014_raw, data2013_raw) %>%
  cs_filter_count(var = Count) -> data2013
```

## Clean-up Enviornment
We can remove the `_raw` objects at this point:

```{r remove-raw}
rm(data2013_raw, data2014_raw, data2015_raw, data2016_raw, data2017_raw, data2018_raw)
```

## Create Single Table
Next, we'll create a single table before we remove individual years. We also subset columns to reduce the footprint of the table.

```{r collapse-all}
bind_rows(data2013, data2014, data2015, data2016, data2017, data2018) %>%
  select(DateOccur, Crime, Description, ILEADSAddress, ILEADSStreet, XCoord, YCoord) %>%
  write_csv(here("data", "clean", "crimes.csv"))
```

### Clean-up Enviornment
We'll remove excess objects again:

```{r remove-years}
rm(data2013, data2014, data2015, data2016, data2017, data2018)
```
